18.821 Project 2

### 2014-10-27 paper/paper.tex
Paper improvements

In this paper we will consider investigate pattern avoidance in permutations of finite sets of positive integers.

\begin{definition}
Two finite sequences $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_n\}$ and $\{b_1,\cdots, b_n\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$. 
\end{definition}

@@

\begin{definition} 
When $n\ge k$, a permutation $\sigma\in S_n$ avoids a permutation $\pi\in S_k$ if $pi$ and $\sigma$ avoid each other when written as sequences in one-line notation. 
\end{definition}

\begin{definition}
For a permutation $\pi$, we write $s_n(\pi$ for the number of permutations in $S_n$ that avoid $\pi$. 
\end{definition}

In this paper we concern ourselves with the sequences $s_n(\pi)$ for various $\pi$. In section \ref{S3} we will consider sequence $s_n(\pi)$ for $\pi\in S_3$, and in section \ref{S4} we will consider sequence $s_n(\pi)$ for $\pi\in S_4$. In section \ref{Tn} we consider a variation on the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of $S_n$.

@@

\subsection{Flipping}

@@

From this we conclude that only permutations of the form described above can avoid $312$.

@@

@@

>>>>

In this paper we will investigate pattern avoidance in permutations of finite sets of positive integers.

Before, beginning our discussion, we present the following definitions.

\begin{definition}
Two finite sequences $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$. 
\end{definition}

@@

\begin{definition} 
When $n\ge k$, a permutation $\sigma\in S_n$ avoids a permutation $\pi\in S_k$ if $\pi$ and $\sigma$ avoid each other when written as sequences in one-line notation. 
\end{definition}

\begin{definition}
For a permutation $\pi$, $s_n(\pi)$ represents the number of permutations in $S_n$ that avoid $\pi$. 
\end{definition}

In this paper we concern ourselves with the sequence of numbers $s_n(\pi)$ for various $\pi$. In section \ref{S3} we will consider sequence $s_n(\pi)$ for $\pi\in S_3$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers. In section \ref{S4} we will consider sequence $s_n(\pi)$ for $\pi\in S_4$. In section \ref{Tn} we consider a variation on the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of $S_n$.

@@

\subsection{Flipping}
Another concept that we will use in this paper is the concept of flipping. We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with elements in the opposite order. More concretely, if $a = a_1, a_2, \dots, a_n$ for some arbitrary $n$, then the flip of the sequence $a$ will be the sequence $b = a_n, a_{n-1}, \dots, a_2, a_1$.

We will talk about how the flip of a sequence is important to this avoidance problem later in the paper.

@@

From this we conclude that all $x < i$ must be to the left of all $y > i$. This information coupled with the fact that we are only considering permutations that end with $i$ complete the claim that only permutations of the form described above can avoid $312$.

@@

It is easy to see that the above proof can be modified slightly to prove that the sequences $s_n(132)$, $s_n(213)$ and $s_n(231)$ are in fact the sequence of catalan numbers as well. In fact, we claim that if $\sigma$ and $\pi$ are both permutations in $S_k$ for some arbitrary $k$, then if $\sigma = flip(\pi)$, $s_n(\pi) = s_n(\sigma)$. This is easy to see since if $a$ is a permutation that avoids $\sigma$, then $flip(a)$ is a permutation that avoids $\pi$ -- since this can be done both ways, it is bijective and we can conclude that the cardinalities of the two sets are equal, or $s_n(\sigma) = s_n(\pi)$.

@@

Studying avoidance of permutations in $T_n$ seems an interesting exercise. In the following section, we consider the sequence of numbers $t_n(\pi)$, defined as the number of permutations in $T_n$ that avoid the permutation $\pi$. In particular, we consider permutations $\pi \in S_3$.

Before stating our conjectures, we present the computed sequences. Note that since the sequence $T_n$ is defined only for even $n$, we only present the even terms of the sequence $t_n$.

$$t_n(123) = 1, 0, 0, 0, 0, \ldots$$
$$t_n(132) = 1, 1, 1, 1, 1, \ldots$$
$$t_n(213) = 1, 2, 4, 8, 16, \ldots$$
$$t_n(231) = 1, 2, 4, 8, 16, \ldots$$
$$t_n(312) = 1, 3, 12, 55, 273, \ldots$$
$$t_n(321) = 1, 3, 12, 55, 273, \ldots$$

It is easy to see why $t_n(123)$ and $t_n(132)$ behave the way they do -- since all odd numbers in $T_n$ are increasing, it is impossible to have a permutation in $T_n$ that avoids $123$ and the permutation consisting of all increasing integers is the only permutation of $T_n$ that avoids $132$.

More interesting is the equivalences between $213$ and $231$, and $312$ and $321$. Proving these equivalences remains future work.
